---
title: "Exploratory Data Analysis (EDA): Station and Delay Analysis"
output: 
    html_document: 
      toc: true
      toc_float: true
      highlight: tango
      theme: simplex
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval=T, echo=F, warning=F, message=F)
```

To understand the nature of the subway delays, the following initial analyses were performed:

* **Summary Statistics:**
    * Calculated the total number of delays and average delay duration at each station.
    * Identified the most common delay causes and their frequency.
* **Visualizations:**
    * **Treemap:** Delay reasons visualized in proportional blocks.
    * **Bar Chart:** Number of delays per subway line and station.
* **Time-Based Analysis:**
    * Created peak (7-9 AM, 3-7 PM) vs non-peak delay summaries.
    * Examined how delay frequencies change by the hour.

```{r}
source("eda.R")
```

## Prelimary Results {.tabset}

### Delays Per Station

```{r}
library(plotly)

p1 <- ggplot(station_delays, aes(x = reorder(Station, Total_Delays), y = Total_Delays, fill = Line, text = paste("Station:", Station, "<br>Delays:", Total_Delays, "<br>Line:", Line))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Number of Delays per Subway Station",
       x = "Subway Station", y = "Number of Delays", fill = "Subway Line") +
  theme_minimal()

ggplotly(p1, tooltip = "text")
```

From this, we can see that Kipling, Bloor, Kennedy, Eglinton, and Finch stations have the highest number of delays. These are all either transfer stations or are terminal stations (with the exception of Eglinton which is one of the busiest stations and the site of a future transfer station). We can also see that more Yonge-University line stations have have a large amount of delays, Kipling and Kennedy, which are are on the BD line have the most and third most delays. We can further look into this by exploring the types of delays that are occurring at stations.

### Reasons For Delay

```{r}
library(highcharter)

delay_causes_treemap <- delay_causes |> 
  mutate(name = `Code Description`,
         value = Count,
         avgDelay = `Average Delay Time`) |> 
  select(name, value, avgDelay)  # Drop extra columns that may cause conflicts

hchart(
  type = "treemap",
  delay_causes_treemap,
  hcaes(x = name, value = value, color = value),
  tooltip = list(
    pointFormat = "<b>{point.name}</b><br>Count: {point.value}<br>Avg Delay: {point.avgDelay:.2f} mins"
  )
) |> 
  hc_title(text = "Top 10 Reasons of Delay") |> 
  hc_colorAxis(stops = color_stops(5, colors = c("indianred1", "indianred4"))) |> 
  hc_chart(backgroundColor = "#fcfcfc")
```

The top reasons for subway delays in Toronto seem to be passenger-related incidents, with "Disorderly Patron" being the most frequent cause, accounting for 887 delays with an average delay of 6.6 minutes. Other notable causes include Passenger Assistance Alarm Activation (652 delays, 4.1 min avg) and OPTO (COMMS) Train Door Monitoring (581 delays, 5.9 min avg).

### Total Delay Time

```{r}
p2 <- ggplot(delay_summary, aes(x = reorder(Code, Total_Delay_Time), y = Total_Delay_Time, fill = Total_Delay_Time, text = paste("Code:", Code, "<br>Description:", `Code Description`, "<br>Total Delay:", Total_Delay_Time, "min"))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Delay Causes by Total Delay Time",
       x = "Delay Code", y = "Total Delay Time (Minutes)", fill = "Total Delay (min)") +
  scale_fill_gradient(low = "indianred1", high = "indianred4") +
  theme_minimal()

ggplotly(p2, tooltip = "text")
```

The top delay causes by total delay time highlight disorderly patrons as the most significant issue, causing over 5,000 minutes of delays throughout the year. Other major contributors include unauthorized people at track level, and injured or ill customers, each accumulating several hundred minutes of delays. Passenger-related incidents, including assaults, alarms, and other disruptions, collectively contribute to substantial downtime. The difference we see with this and the last visualization is that a few of the most commonly occurring delays, such as the passenger alarm being activated and unsanitary vehicle are not in the top 10 of total delay time since the average time for those delays is much less.

### Delays Time of Day

```{r}
p4
kable(peak_summary, caption = "Delays During Peak vs Non-Peak Hours")
```

Figure 4 shows the occurrence of delays distributed by the time of day at which they occur. The peak hours are based on the official TTC rush hour times when train service is increased. There is a significant jump in the number of delays during the rush hours with the highest delays occurring around 7am and 5pm. What is interesting is that while the number of delays is much higher during peak hours, the average time of each delay is almost the same (they only differ by about 15 seconds).

## Summary
This analysis of TTC subway delays has provided several key insights into delay causes, timing patterns, and station-specific trends. The most frequent delay and the delay that has wasted the most amount of time this year is **Disorderly Patrons**, accounting for 3405 recorded delays and taking a total of 8517 hours, much more than any other delay type. Other common delay types include passenger assistance alarms being activated, train door monitoring issues, and medical emergencies. However, when examining total delay time, disruptions such as unauthorized people at track level and injured or ill customer on train incidents accumulate significantly more system downtime, despite occurring less frequently.

In terms of time-based patters, delays are more frequent during peak hours (7-9 AM, 3-7 PM) when the TTC is running more trains, with a noticeable spike between 4 and 6 PM. Interestingly, while peak hours have more delays, the average duration of each delay is similar to non-peak periods, suggestions that the nature of delays remains relatively consistent through the day. Additionally, certain stations such as Kipling, Bloor, Kennedy and Finch experience the highest number of delays, many of which are near major transfer or terminal stations.
